\Large \bf 6.829 Final Project: A comparison of bit-rate selection algorithms

### 2013-05-16 paper/paper.tex
spell checked

%\usepackage{endnotes}
\begin{document}

%don't want date printed

@@

There are three main classes of bit rate adaptiation protocols: frame-based, SNR-based, and cross-layer protocols. Frame-based protocols measure the fraction of sucessfully received protocols. SNR protocols make decisions based on the estimated Signal to Noise ratio. Cross-layer protocols use SoftPHY data from the physical layer. The most commonly implemented protocols on today's networks are frame based, because SNR protocols perform poorly~\cite{samplerate}, and cross-layer protocols cannot be deployed on current networks because they violate the network layering abstration.

The two most popular fram-based protocols are SampleRate and Minstrel. Both were implemented in the MadWifi drivers for Linux wirless, and today Minstrel is the default bit-rate selection algorithm for all wireless drivers on Linux. Our research used a trace-based approach to analyze the performance of these two algorithms. We also created a modified version of Minstrel that provides significant throughput gains over the vanilla Minstrel implementation.

@@

SampleRate has three main functions: \texttt{ApplyRate()}, which returns a the bit-rate to send a packet at, \texttt{ProcessFeedback()}, which updates the statistics for a bitrate after an attempted packet send,  and \texttt{RemoveStaleResults()}, which removes all statistics older than 10 seconds.

@@

Minstrel utilizes the multi-rate retry chain (mrr), which is an array that tells the card which rates to try before reporting a failure. The retry chain makes failures extremely unlikely. Minstrel has a mrr of [\emph{highest throughput, next highest throughput, best probability, lowest base rate}] for most packets. In essence, Minstrel tells the card to send at the highest throughput rates but to fall back to reliable rates upon failure. The throughput and success probability are calculated for each rate every 100ms. 
To keep an accurate estimate of throughput and probability for each rate, Minstrel sends sample frames ten percent of the time. If the randomly chosen rate has a higher lossless throughput than the current optimal rate, the mrr is as follows: [\emph{random, highest throughput, best probability, lowest base rate}]. If the random rate is slower than the optimal rate, it is moved further back in the retry chain: [\emph{best throughput, random, best probability, lowest base rate}]. This ensures that Minstrel never samples rates worse than the current optimal rate unless the optimal rate experiences a failure. See table ~\ref{table:1} for an overview of the mrr's used in Minsterl.

@@

The probability $p$ of a successful transmission is calculated from statistics collected by the \texttt{ProcessFeedback()} method, and is done using an exponential weighted moving average, or EWMA. The EWMA creates a weighted average that weighs recent data more heavily than old data. The weighting for old data decreases exponentially. An EWMA is beneficial because it ensures that a sudden degredation in link quality will create a rapid response in the probabilities due to the increased weight of recent data. Minstrel will waste less time sending packets at a rate that no longer works, as compared to a simple probability calculation.

@@

Our initial plan was to use the algorithms as implemented in the MadWifi drivers for the Atheros chipset. However, MadWifi is depricated, and only runs on quite old network cards. In addition to the lack of compatible hardware available to us, MadWifi is not an accurate reflection of what a modern user would use. We looked into the ath5k and ath9k projects, which were created to replace MadWifi. Both of these did not port the old bit-rate selection system from MadWifi and instead opted to use the linux kernel's rate selection software, which only offers Minstrel. Porting SampleRate over to the new drivers would not have been trivial, as there are vast differences in the assumed interfaces.

@@

We implemented SampleRate as outlined in John Bicket's Master's thesis~\cite{samplerate}. This implementation is slightly different from how it was implemented in the MadWifi kernel driver. The primary difference is that the kernel implementatiion uses a EWMA instead of a 10 second window. The thesis implmenentation throws out old rate data and begins anew every 10 seconds. It would be worth looking at the performance of EWMA-based SampleRate in the future.

@@

Minstrel and SampleRate perform similarly, usually within 3 Mbps of each other. In especially noisy situations, such as when the client is moving or when the client is around the corner from the access point, Minstrel consistently outperforms SampleRate. Figure~\ref{figure:3} shows the results of a single run of both Minstrel and SampleRate over all the different types of traces we collected. This is consisten with the literature---the authors of Minstrel noted that they focused on making Minstrel robust to poor conditions~\cite{minstrel}, which means losing potential througput gains in stable situations. Conversely, John Bicket noted that SampleRate performs better in stable situations than noisy or highly varying conditions~\cite{samplerate}. We were slightly surprised to see how significantly SampleRate outperformed Minstrel in certain situations, though, with throughput being up to 67\% higher.

In Figure~\ref{figure:2} we show the performance of constant bitrates on the \texttt{clear\_1.dat} trace, which is a 35 second trace in clear view of the access point. The optimal rate here was 18 Mbps, which achieved an average throughput of about 15 Mbps. The next highest bitrate, 24 Mbps, shows a large plummet in performance, down to a throughput of 1.17 Mbps. This steep drop is charateristic, and is also talked about in the SampleRate thesis~\cite{samplerate}. It demonstrates just how important bit rate selection is---the optimal bit rate has typically much higher throughput than the next-best rate, but if the algorithm strays too high the results can be devistating.

@@

The histograms of which rates the algorithms deided to send at provided some useful insight into the behavior. Table~\ref{table:2} is a histogram of a 35 second corner trace. In this situation Minstrel outperforms SampleRate. The constant rate throughputs demonstrate that 11, 12 and 18 Mbps were the optimal rates to send at, and that no packets were sucessful at any higher rates. SampleRate does not use these rates except for probe packets, which happens once every ten seconds. Minstrel does not have this measure, however, spends a lot of time sending probe packets at these high rates that happen to never be successful. The only reason low rates have fewer probe packets is because they are second in the retry chain. Despite the fact that SampleRate spends less time probing high rates, it is less successful. This is because Minstrel send the most packets at the three most successful bitrates. SampleRate, however, spends too much time sampling low bit rates. Additionally, it spends a lot of time sending at 5.5 Mbps. TODO: why does SampleRate spend time at 5.5 Mbps??? We suspect that this might be because the equation SampleRate uses to estimate average transmission time needs to be tweaked.

Our examination of the histograms showed us that Minstrel spends a lot of time sampling at high rates. In the specific case we examined, they were all probe packets. But sometimes Minstrel would choose 54Mbps as the rate with the best throughput, which made little sense. A closer investigation revealed that Minstrel would sometimes luck out and get one success sending a probe at 54Mbps. The probability of success would be 1, so then 54Mbps would be chosen as the rate with the best throughput. This is unfortunate, but the real harm came in the next window: nearly all of the packets at 54Mbps would fail, but Minstrel was still considering 54 Mbps to be the best rate! This is because the EWMA did not account for the fact that the second block having sent hunreds more packets at 54 Mbps than the first. Thus the probability of success would drop from 1 to about 0.70, which is still high enough for the throughput equation to rank 54 Mbps highly. In essense, the probabilities were not declining fast enough after a large number of failures.

@@

Very little analysis has been done of the performance of bit rate selection algorithms on 802.11n networks. 802.11n introduces many new rates, as well as multiple-input/multiple output, or MIMO, capabilities. There is an 802.11n version of Minstrel, but there is no implmentation of SampleRate.

@@

Hari Balakrishnan suggested that we use a broadcast approach when collecting our traces, instead of link layer acknowledgements. This would entail having multiple computers listening at the broadcast address, and compiling successes and failures from that information. This approach would free us from certain idiosyncracies of reading data from the driver, and lead to possibly more accurate traces. We were unable to implement this due to a lack of time and ability to acquire compatible hardware.

@@

Minstrel and SampleRate perform similarly. The results suggest that SampleRate tends to do slightly better, except in very lossy links such as the client being far awat from the AP or actively moving. Minproved, our improved version of Minstrel, solidly outperforms both Minstrel and SampleRate.

@@

\section{Acknowledgements}
We would like to thank Jonathan Perry and Hari Balakrishnan for their valuable guidance, as well as Derek Smithies for answering our questions about Minstrel.

>>>>

%\usepackage{endnotes}
\begin{document}
%don't want date printed

@@

There are three main classes of bit rate adaptation protocols: frame-based, SNR-based, and cross-layer protocols. Frame-based protocols measure the fraction of successfully received packets. SNR protocols make decisions based on the estimated Signal to Noise ratio. Cross-layer protocols use SoftPHY data from the physical layer. The most commonly implemented protocols on today's networks are frame based, because SNR protocols perform poorly~\cite{samplerate}, and cross-layer protocols cannot be deployed on current networks because they violate the network layering abstraction.

The two most popular frame-based protocols are SampleRate and Minstrel. Both were implemented in the MadWifi drivers for Linux wireless, and today Minstrel is the default bit-rate selection algorithm for all wireless drivers on Linux. Our research used a trace-based approach to analyze the performance of these two algorithms. We also created a modified version of Minstrel that provides significant throughput gains over the vanilla Minstrel implementation.

@@

SampleRate has three main functions: \texttt{ApplyRate()}, which returns a the bit-rate to send a packet at, \texttt{ProcessFeedback()}, which updates the statistics for a bit rate after an attempted packet send,  and \texttt{RemoveStaleResults()}, which removes all statistics older than 10 seconds.

@@

Minstrel utilizes the multi-rate retry chain (mrr), which is an array that tells the card which rates to try before reporting a failure. The retry chain makes failures extremely unlikely. Minstrel has a mrr of [\emph{highest throughput, next highest throughput, best probability, lowest base rate}] for most packets. In essence, Minstrel tells the card to send at the highest throughput rates but to fall back to reliable rates upon failure. The throughput and success probability are calculated for each rate every 100ms. 
To keep an accurate estimate of throughput and probability for each rate, Minstrel sends sample frames ten percent of the time. If the randomly chosen rate has a higher lossless throughput than the current optimal rate, the mrr is as follows: [\emph{random, highest throughput, best probability, lowest base rate}]. If the random rate is slower than the optimal rate, it is moved further back in the retry chain: [\emph{best throughput, random, best probability, lowest base rate}]. This ensures that Minstrel never samples rates worse than the current optimal rate unless the optimal rate experiences a failure. See table ~\ref{table:1} for an overview of the mrr's used in Minstrel.

@@

The probability $p$ of a successful transmission is calculated from statistics collected by the \texttt{ProcessFeedback()} method, and is done using an exponential weighted moving average, or EWMA. The EWMA creates a weighted average that weighs recent data more heavily than old data. The weighting for old data decreases exponentially. An EWMA is beneficial because it ensures that a sudden degradation in link quality will create a rapid response in the probabilities due to the increased weight of recent data. Minstrel will waste less time sending packets at a rate that no longer works, as compared to a simple probability calculation.

@@

Our initial plan was to use the algorithms as implemented in the MadWifi drivers for the Atheros chipset. However, MadWifi is deprecated, and only runs on quite old network cards. In addition to the lack of compatible hardware available to us, MadWifi is not an accurate reflection of what a modern user would use. We looked into the ath5k and ath9k projects, which were created to replace MadWifi. Both of these did not port the old bit-rate selection system from MadWifi and instead opted to use the Linux kernel's rate selection software, which only offers Minstrel. Porting SampleRate over to the new drivers would not have been trivial, as there are vast differences in the assumed interfaces.

@@

We implemented SampleRate as outlined in John Bicket's Master's thesis~\cite{samplerate}. This implementation is slightly different from how it was implemented in the MadWifi kernel driver. The primary difference is that the kernel implementation uses a EWMA instead of a 10 second window. The thesis implementation throws out old rate data and begins anew every 10 seconds. It would be worth looking at the performance of EWMA-based SampleRate in the future.

@@

Minstrel and SampleRate perform similarly, usually within 3 Mbps of each other. In especially noisy situations, such as when the client is moving or when the client is around the corner from the access point, Minstrel consistently outperforms SampleRate. Figure~\ref{figure:3} shows the results of a single run of both Minstrel and SampleRate over all the different types of traces we collected. This is consistent with the literature---the authors of Minstrel noted that they focused on making Minstrel robust to poor conditions~\cite{minstrel}, which means losing potential throughput gains in stable situations. Conversely, John Bicket noted that SampleRate performs better in stable situations than noisy or highly varying conditions~\cite{samplerate}. We were slightly surprised to see how significantly SampleRate outperformed Minstrel in certain situations, though, with throughput being up to 67\% higher.

In Figure~\ref{figure:2} we show the performance of constant bit rates on the \texttt{clear\_1.dat} trace, which is a 35 second trace in clear view of the access point. The optimal rate here was 18 Mbps, which achieved an average throughput of about 15 Mbps. The next highest bitrate, 24 Mbps, shows a large plummet in performance, down to a throughput of 1.17 Mbps. This steep drop is characteristic, and is also talked about in the SampleRate thesis~\cite{samplerate}. It demonstrates just how important bit rate selection is---the optimal bit rate has typically much higher throughput than the next-best rate, but if the algorithm strays too high the results can be devastating.

@@

The histograms of which rates the algorithms decided to send at provided some useful insight into the behavior. Table~\ref{table:2} is a histogram of a 35 second corner trace. In this situation Minstrel outperforms SampleRate. The constant rate throughputs demonstrate that 11, 12 and 18 Mbps were the optimal rates to send at, and that no packets were successful at any higher rates. SampleRate does not use these rates except for probe packets, which happens once every ten seconds. Minstrel does not have this measure, however, spends a lot of time sending probe packets at these high rates that happen to never be successful. The only reason low rates have fewer probe packets is because they are second in the retry chain. Despite the fact that SampleRate spends less time probing high rates, it is less successful. This is because Minstrel send the most packets at the three most successful bitrates. SampleRate, however, spends too much time sampling low bit rates. Additionally, it spends a lot of time sending at 5.5 Mbps. TODO: why does SampleRate spend time at 5.5 Mbps??? We suspect that this might be because the equation SampleRate uses to estimate average transmission time needs to be tweaked.

Our examination of the histograms showed us that Minstrel spends a lot of time sampling at high rates. In the specific case we examined, they were all probe packets. But sometimes Minstrel would choose 54 Mbps as the rate with the best throughput, which made little sense. A closer investigation revealed that Minstrel would sometimes luck out and get one success sending a probe at 54 Mbps. The probability of success would be 1, so then 54 Mbps would be chosen as the rate with the best throughput. This is unfortunate, but the real harm came in the next window: nearly all of the packets at 54 Mbps would fail, but Minstrel was still considering 54 Mbps to be the best rate! This is because the EWMA did not account for the fact that the second block having sent hundreds more packets at 54 Mbps than the first. Thus the probability of success would drop from 1 to about 0.70, which is still high enough for the throughput equation to rank 54 Mbps highly. In essence, the probabilities were not declining fast enough after a large number of failures.

@@

Very little analysis has been done of the performance of bit rate selection algorithms on 802.11n networks. 802.11n introduces many new rates, as well as multiple-input/multiple output, or MIMO, capabilities. There is an 802.11n version of Minstrel, but there is no implementation of SampleRate.

@@

Hari Balakrishnan suggested that we use a broadcast approach when collecting our traces, instead of link layer acknowledgments. This would entail having multiple computers listening at the broadcast address, and compiling successes and failures from that information. This approach would free us from certain idiosyncrasies of reading data from the driver, and lead to possibly more accurate traces. We were unable to implement this due to a lack of time and ability to acquire compatible hardware.

@@

Minstrel and SampleRate perform similarly. The results suggest that SampleRate tends to do slightly better, except in very lossy links such as the client being far away from the AP or actively moving. Minproved, our improved version of Minstrel, solidly outperforms both Minstrel and SampleRate.

@@

\section{Acknowledgments}
We would like to thank Jonathan Perry and Hari Balakrishnan for their valuable guidance, as well as Derek Smithies for answering our questions about Minstrel.
